---
title: "Logistic Model"
author: "Elise Longenecker & Joanna Schroeder"
date: "2025-05-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load the data and relevant packages

```{r}
data <- read.csv("C:/Users/elise/OneDrive/Documents/ML-AI_Algorithms_I/Final/actors_features.csv")
library(dplyr)
library(lubridate)
library(stringr)
library(purrr)
```

### Filtering data to only include actors who received an Oscar nomination

```{r}
data <- data %>%
  filter(nominee == 1) %>%
  mutate(min_year = map_dbl(nominated_years, function(s) {
    s %>%
      str_remove_all("\\[|\\]") %>%
      str_split(",\\s*") %>%
      unlist() %>%
      as.numeric() %>%
      min(na.rm = TRUE)
  })) %>%
  mutate(age_at_nomination = as.numeric(difftime(min_year, birth_year)))
```

### Creating a binary variable for if the person was nominated as a leading actor

```{r}
keyword1 <- "LEADING"
keyword2 <- "BOTH"
data$leading <- ifelse(
  grepl(keyword1, data$oscar_type, ignore.case = TRUE) |
    grepl(keyword2, data$oscar_type, ignore.case = TRUE),
  1, 0
)
```

### Splitting the data into 80% train, 20% test

```{r}
train_indices <- sample(1:nrow(data), size = 0.8 * nrow(data))
train <- data[train_indices, ]
test <- data[-train_indices, ]
```

### Logistic model for the likelihood of a person winning the Oscar, given their age, gender, leading or supporting nomination, whether or not they are American, and their popularity score.

```{r}
logistic_winner <- glm(winner ~ age_at_nomination + gender + leading + american + Popularity, data = train, family = binomial)
summary(logistic_winner)
```

### Predicting the test data using the model

```{r}
test$predicted_prob_winner <- predict(logistic_winner, newdata = test, type = "response")
```

